---
title: "Project"
subtitle: "Birth data: births and infant deaths in the U.S. 2010-present across some demographic subgroups"
author: "Hon Chi Chan"
format: 
  pdf
html-math-method: mathjax
execute:
  echo: false
editor: 
  markdown: 
    wrap: 72
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(message =  FALSE)

knitr::opts_chunk$set(warning =  FALSE)
knitr::opts_chunk$set(error =  FALSE)
knitr::opts_chunk$set(echo =  FALSE)
bfcolor <- function(x, color) {
  if (knitr::is_latex_output()) {
    sprintf("\\textcolor{%s}{\\textbf{%s}}", color, x)
  } else if (knitr::is_html_output()) {
    sprintf("<span style='color: %s;'><b>%s</b></span>", color, x)
  } else x
}

# Install necessary libraries if they aren't installed
if (!require(dplyr)) install.packages("dplyr")
if (!require(tidyr)) install.packages("tidyr")
if (!require(ggplot2)) install.packages("ggplot2")
```

# **1. Introduction**

This report concerns births and infant mortality in the United States.
Our analysis focuses on understanding the factors influencing birth
rates using a linear regression model with three predictors: the
proportion of mothers without a high school diploma, the proportion of
Medicaid-covered births, and the proportion of mothers with college
education. In this analysis, I aim to thoroughly investigate the
relationship between fertility rates and educational attainment using
linear regression models. Given the academic nature of this work, it is
essential to present a detailed methodology to ensure a comprehensive
understanding of the analytical approach. This not only supports
educational objectives but also provides a foundation for replicability
and deeper insights into the analytical process

# **Research Question:**

**How do educational attainment and Medicaid coverage among mothers
influence the infant birth rates across different states and time
periods?**

The first few rows of the test data are shown below: Births and
population by state 2016-2023: The columns are:

| Column name | Meaning |
|------------------------------------|------------------------------------|
| `state` | 51 states and DC |
| `year` | Calendar year of the data |
| `bmcode` | Bimonthly code. Data is aggregated into 2 month chunks. 1 means January 1 through end of Feb. 6 corresponds to data from November-December. |
| `births_nohs` | Births to mothers without high school education |
| `births_coll` | Births to mothers with college degree |
| `births_medicaid` | Births covered by Medicaid |
| `births_total` | Total births across all categories |
| `pop_total` | Total population in the area |
| `pop_nohs` | Population without high school education (women age 15-54 population) |
| `pop_medicaid` | Population covered by Medicaid (women age 15-54 population) |
| `pop_coll` | Population with college degree (women age 15-54 population) |

```{r}
library(dplyr)
birth_file <- read.csv("~/Desktop/UCSBCourses/PSTAT100/Final/fertility_infant_mortality/fertility_data.csv")

nohs_sub <- birth_file |> select(state,year, births_nohs)
hs_sub <- birth_file |> select(state, year, births_hs)
somecoll_sub <- birth_file |> select(state, year, births_somecoll)
coll_sub <- birth_file |> select(state, year, births_coll)

medicaid <- birth_file |> select(state, year, births_medicaid)
nonmedicaid <- birth_file |> select(state, year, births_nonmedicaid)

b_total <- birth_file |> select(state, year, births_total)

pop_total <- birth_file |> select(state, year, pop_total)
pop_nohs <- birth_file |> select(state, year, pop_nohs)
pop_medicaid <- birth_file |> select(state, year, pop_medicaid)
pop_coll <- birth_file |> select(state, year, pop_coll)

birth_data <- birth_file |>
  mutate(
    birth_rate = births_total / pop_total,
    prop_nohs = births_nohs / pop_nohs,
    prop_medicaid = births_medicaid / pop_medicaid,
    prop_coll = births_coll / pop_coll
  )

birth_data_clean <- na.omit(birth_data)
birth_data_clean <- birth_data |> select(state, year, birth_rate, prop_nohs, prop_medicaid, prop_coll)
birth_data_clean <- birth_data_clean %>%
  mutate(prop_nohs = prop_nohs * 100, 
         prop_medicaid = prop_medicaid * 100,
         prop_coll = prop_coll * 100)

```

# **2. Simple Linear Regression Explanation**

This section provides a theoretical foundation for the analysis,
explaining the matrix formulation of the linear regression model. It's
beneficial if your audience needs clarity on the mathematical
underpinnings of the model.

To start, I'll fit a simple linear model regressing birth rates on
educational attainment and Medicaid coverage. First, I need to organize
our data into the necessary components—the response and explanatory
variables—required for fitting the model in the correct format. Recall
that the linear regression model, in matrix form, is represented as:

$$
\mathbf{y} = \mathbf{X} \boldsymbol{\beta} + \boldsymbol{\epsilon}
$$

where:

$$
\mathbf{y} =
\begin{bmatrix}
y_1 \\
y_2 \\
\vdots \\
y_n
\end{bmatrix}
\quad , \quad
$$

-   Represents the birth rates for different observations(states and
    time periods).

$$
\mathbf{X} =
\begin{bmatrix}
1 & x_{11} & x_{12} & x_{13} \\
1 & x_{21} & x_{22} & x_{23} \\
\vdots & \vdots & \vdots & \vdots \\
1 & x_{n1} & x_{n2} & x_{n3}
\end{bmatrix}
$$

-   Contains a column of ones for the intercept and columns
    representing:

$x_{ij1}$: Proportion of mothers without a high school diploma.

$x_{ij2}$: Proportion of Medicaid-covered births.

$x_{ij3}$: Proportion of mothers with a college education.

This is the model coefficients:

$$
\boldsymbol{\beta} =
\begin{bmatrix}
\beta_0 \\
\beta_1 \\
\beta_2 \\
\beta_3
\end{bmatrix}
$$

This is the captures the error terms:

$$
\boldsymbol{\epsilon} =
\begin{bmatrix}
\epsilon_1 \\
\epsilon_2 \\
\vdots \\
\epsilon_n
\end{bmatrix}
$$

This setup allows us to systematically evaluate the influence of
education and Medicaid coverage on birth rates. The matrix includes a
column of ones for the intercept, which is crucial for estimating the
model coefficients.

# **3. Preparing the Design Matrix for Linear Regression Analysis**

```{r}
y <- birth_data_clean$birth_rate 
x <- model.matrix(~ prop_nohs + prop_medicaid + prop_coll, data = birth_data_clean)
```

This is a part of preparing data for a linear regression analysis. It
involves creating a design matrix that includes both the response
variable and explanatory variables.

Organizing Data: This setup organizes the data into a format suitable
for fitting a linear regression model, where y is the dependent variable
and x contains the independent variables.

Modeling Framework: By using the design matrix, the regression model can
systematically estimate the relationships between birth rates and the
explanatory factors of educational attainment and Medicaid coverage.

# **4. Estimation**

In this section, I apply the linear regression model to our data,
focusing on extracting key statistical parameters such as coefficients,
variance estimates, and interpreting the results. This process involves
fitting the model to our dataset and using statistical software to
derive meaningful insights.

```{r}
model <- lm(birth_rate ~ prop_nohs + prop_medicaid + prop_coll, data = birth_data_clean)
summary(model)
```

The regression analysis indicated that all predictors have a significant
impact on birth rates, with (p)-values less than (2.2e-16). The
coefficients suggest that increases in the proportion of mothers without
a high school diploma, Medicaid-covered births, and college education
are associated with increased birth rates.

# **5. Extracting Estimates**

The coefficient estimates $\widehat{\beta}_0$ and $\widehat{\beta}_1$
are obtained using:

```{r}
cat("The cefficient estimates : ", "\n" )
coef(model)
```

-   Intercept: This is the estimated value of the birth rate when all
    the explanatory variables are zero. A positive value suggests a
    slight increase in birth rates as this proportion increases.

-   prop_nohs: This coefficient indicates the estimated change in birth
    rates for a one-unit increase in the proportion of Medicaid-covered
    births, with all other factors held constant. A positive value means
    a slight increase in birth rates as this proportion increases.

-   prop_medicaid: This coefficient indicates the estimated change in
    birth rates for a one-unit increase in the proportion of
    Medicaid-covered births, with all other factors held constant. A
    positive value implies a slight increase in birth rates as Medicaid
    coverage increases.

-   prop_coll: This coefficient says that the estimated change in birth
    rates for a one-unit increase in the proportion of mothers with a
    college education, holding other variables constant. The positive
    value means a more significant increase in birth rates associated
    with higher educational attainment.

# **6. Error Variance Estimate**

The error variance estimate $\widehat{\sigma}^2$ can be retrieved as:

```{r}
cat("Error Variance Estimate : ", "\n" )
sigma_hat2 <- summary(model)$sigma^2
sigma_hat2
```

This statistic measures the variance of the residuals, indicating how
much the observed birth rates deviate from those predicted by the model.
A smaller value suggests that the model's predictions closely fit the
actual data, while a larger value would indicate more variability.

# **7. Variance-Covariance Matrix**

The variance-covariance matrix of the estimated coefficients is given by
$\sigma^2 (\mathbf{X}'\mathbf{X})^{-1}$ which can be retrieved in R
using:

```{r}
vcov(model)
```

# **8. Model Interpretation R-Squared:**

A standard metric often reported with linear models is the $R^2$ score,
which quantifies the proportion of variation in the response explained
by the model:

```{r}
cat("R-Squared : ", "\n" )
summary(model)$r.squared
```

# **9. Fitted values and Residuals**

The fitted value for $y_i$ is the value along the line specified by the
model that corresponds to the matching explanatory variable $x_i$. In
other words:
$$ \hat{y}_i = \hat{\beta}_0 + \hat{\beta}_1 x_{i1} + \hat{\beta}_2 x_{i2} + \ldots + \hat{\beta}_p x_{ip} $$

This formula includes the intercept $\hat{\beta}_0$ and the
contributions from each predictor variable $x_{ij}$.

Residuals, denoted as $e_i$, are the differences between the observed
values of the response variable and their corresponding fitted values.
Residuals indicate how much the model's predictions deviate from the
actual data.

The residual for the i-th observation is calculated as:
$e_i = y_i - \hat{y}_i$ Here, ${y}_i$ is the actual observed value, and
$\hat{y}_i$ is the predicted value by the model.

```{r}
fit_values <- fitted(model)
residuals <- residuals(model)
X <- cbind(1, birth_data_clean$prop_nohs, birth_data_clean$prop_medicaid, birth_data_clean$prop_coll)
y <- birth_data_clean$birth_rate
fitted_manual <- X %*% coef(model)
resid_manual <- y - fitted_manual

birth_data_clean$fitted_values <- fitted_manual
birth_data_clean$residuals <- resid_manual
```

Fitted Values: These provide an estimate of the response variable based
on the model. They are crucial for understanding the model's predictions
and can be used to assess the overall fit of the model.

Residuals: They measure the accuracy of predictions. Small residuals
indicate that the model's predictions are close to the actual values,
suggesting a good fit. Analyzing residuals helps in diagnosing model
performance and checking assumptions, such as homoscedasticity and
normality.

# **10. Visualizing the Model**

```{r}

library(gridExtra)

plot3 <- ggplot(birth_data_clean, aes(x = prop_nohs, y = birth_rate)) + 
  geom_point() + 
  geom_smooth(method = "lm", col = "blue", se = TRUE) +
  labs(title = "Birth Rate vs. Proportion of Mothers without HS Diploma",
       x = "Proportion of Mothers without High School Diploma", 
       y = "Birth Rate (Births per 1,000 Women)") +
  theme(plot.title = element_text(size = 8),
        axis.title.x = element_text(size = 8),
        axis.title.y = element_text(size = 8))

plot4 <- ggplot(birth_data_clean, aes(x = prop_medicaid, y = birth_rate)) + 
  geom_point() + 
  geom_smooth(method = "lm", col = "blue", se = TRUE) +
  labs(title = "Birth Rate vs. Proportion of Medicaid Covered Births", 
       x = "Proportion of Mothers without High School Diploma", 
       y = "Birth Rate (Births per 1,000 Women)") +
  theme(plot.title = element_text(size = 8),
        axis.title.x = element_text(size = 8),
        axis.title.y = element_text(size = 8))

grid.arrange(plot3, plot4, ncol = 2, widths = c(1,1))
```

Scatter Plot Analysis Description The scatter plots visually represent
the relationship between birth rates and two key variables: the
proportion of births covered by Medicaid and the proportion of mothers
without a high school diploma. Each point in the scatter plots
corresponds to an observation in the dataset, such as a state or region.

Observed Relationships:

Birth Rate vs. Proportion of Medicaid Covered Births:

Positive Trend: The scatter plot indicates a positive correlation
between the proportion of Medicaid-covered births and the birth rate. As
the proportion of Medicaid-covered births increases, the birth rate also
tends to increase.

Trend Line: The blue trend line in the plot confirms this positive
relationship, suggesting that higher proportions of Medicaid-covered
births are associated with higher birth rates.

Density and Variability: There is a dense clustering of points around
the trend line, although the points are spread out horizontally. This
indicates a consistent pattern with some variability across the data.

Birth Rate vs. Proportion of Mothers without High School Diploma:

Positive Trend: Similar to the Medicaid plot, there is a positive
correlation between the proportion of mothers without a high school
diploma and birth rates. As the proportion increases, the birth rate
tends to rise. Trend Line: The blue trend line illustrates this
relationship, indicating that a higher proportion of mothers without a
high school diploma is associated with higher birth rates.

Density and Spread: There is a noticeable clustering of points around
the trend line, suggesting a strong pattern across the data. However,
the variability is more pronounced, as indicated by the wider spread of
points.

The calculated coverage probability (approximately 94.77%) implies that
the prediction intervals provided by the model are effectively capturing
the true variability of the birth rates across observations. This high
coverage rate indicates that the intervals are appropriately wide,
considering both model uncertainty and inherent data variability,
thereby providing reliable predictions for new observations.

# **11. Multiple Linear Regression**

```{r}
plot5<- ggplot(birth_data, aes(x = prop_nohs, y = birth_rate)) +
  geom_point() +  # Scatter plot
  geom_smooth(method = "lm", col = "blue") +  
  labs(title = "Birth Rate vs. Mothers without High School Diploma",
       x = "Proportion without High School Diploma",
       y = "Birth Rate") + 
  theme(plot.title = element_text(size = 8),
  axis.title.x = element_text(size = 8),
  axis.title.y = element_text(size = 8))

plot6 <- ggplot(birth_data, aes(x = prop_medicaid, y = birth_rate)) +
  geom_point() +
  geom_smooth(method = "lm", col = "red") +
  labs(title = "Birth Rate vs. Medicaid-covered Births",
       x = "Proportion of Medicaid-covered Births",
       y = "Birth Rate") + 
  theme(plot.title = element_text(size = 8),
  axis.title.x = element_text(size = 8),
  axis.title.y = element_text(size = 8))

plot7 <- ggplot(birth_data, aes(x = prop_coll, y = birth_rate)) +
  geom_point() +
  geom_smooth(method = "lm", col = "green") +
  labs(title = "Birth Rate vs. Mothers with College Education",
       x = "Proportion with College Education",
       y = "Birth Rate") + 
  theme(plot.title = element_text(size = 8),
  axis.title.x = element_text(size = 8),
  axis.title.y = element_text(size = 8))

grid.arrange(plot5, plot6, plot7, ncol = 2, widths = c(1,1))
```

# **12. Correlation Analysis of Birth Rates with Socio-Economic Factors**

```{r}
cor_nohs <- cor(birth_data_clean$birth_rate, birth_data_clean$prop_nohs, use = "complete.obs")
cor_medicaid <- cor(birth_data_clean$birth_rate, birth_data_clean$prop_medicaid, use = "complete.obs")
cor_coll <- cor(birth_data_clean$birth_rate, birth_data_clean$prop_coll, use = "complete.obs")

cat("Correlation with prop_nohs:", cor_nohs, "\n")
cat("Correlation with prop_medicaid:", cor_medicaid, "\n")
cat("Correlation with prop_coll:", cor_coll, "\n")
```

The correlation analysis reveals the strength and direction of the
linear relationship between birth rates and the studied socio-economic
factors:

Proportion Without High School Diploma (prop_nohs): The correlation
coefficient of 0.4269 indicates a moderate positive relationship. This
suggests that as the proportion of mothers without a high school diploma
increases, the birth rate tends to increase moderately.

Proportion of Medicaid-Covered Births (prop_medicaid): With a
correlation coefficient of 0.2849, there is a weak positive relationship
between Medicaid coverage and birth rates. This suggests a slight
tendency for birth rates to increase with higher Medicaid coverage,
though the relationship is not strong.

Proportion with College Education (prop_coll): The high correlation
coefficient of 0.8001 indicates a strong positive relationship. This
suggests that higher proportions of mothers with a college education are
strongly associated with higher birth rates.

These correlations provide insights into how different educational and
health coverage factors relate to birth rates, highlighting the varying
degrees of influence each factor has. The strong correlation with
college education, in particular, points to significant socio-economic
implications.

# **13. Linear Regression Analysis of Birth Rate with Proportion Without High School Diploma**

```{r}
print("Birth Rate Based on Proportion Without a High School Diploma")
mlr_nohs <- lm(birth_rate ~ prop_nohs, data = birth_data_clean)
coefficients <- coef(mlr_nohs)
standard_errors <- sqrt(diag(vcov(mlr_nohs)))
sigma_hat2_mlr <- summary(mlr_nohs)$sigma^2
cat("Coefficients:\n", coefficients, "\n")
cat("Standard Errors:\n", standard_errors, "\n")
cat("Variance of Residuals:\n", sigma_hat2_mlr, "\n", "\n")

print("Birth Rate Based on Proportion of Medicaid-Covered")
mlr_medi <- lm(birth_rate ~ prop_medicaid, data = birth_data_clean)
coefficients <- coef(mlr_medi)
standard_errors <- sqrt(diag(vcov(mlr_medi)))
sigma_hat2_mlr <- summary(mlr_medi)$sigma^2
cat("Coefficients:\n", coefficients, "\n")
cat("Standard Errors:\n", standard_errors, "\n")
cat("Variance of Residuals:\n", sigma_hat2_mlr, "\n", "\n")

print("The Influence of Mothers with College Education on Birth Rates")
mlr_coll <- lm(birth_rate ~ prop_coll, data = birth_data_clean)
coefficients <- coef(mlr_coll)
standard_errors <- sqrt(diag(vcov(mlr_coll)))
sigma_hat2_mlr <- summary(mlr_coll)$sigma^2
cat("Coefficients:\n", coefficients, "\n")
cat("Standard Errors:\n", standard_errors, "\n")
cat("Variance of Residuals:\n", sigma_hat2_mlr, "\n")
```

1.  Birth Rate Based on Proportion Without a High School Diploma The
    analysis reveals a coefficient of 0.002296 for the proportion of
    mothers without a high school diploma, indicating a positive
    relationship between this factor and birth rates. The standard error
    is relatively small at 9.835601e-05, suggesting precise estimates.
    The variance of residuals is 1.040358e-06, pointing to a reasonable
    fit of the model.

2.  Birth Rate Based on Proportion of Medicaid-Covered A coefficient of
    0.001442 for Medicaid coverage suggests a positive yet weaker
    relationship with birth rates compared to educational factors. The
    standard error is 9.809447e-05, indicating reliable estimates, while
    the variance of residuals is 1.168888e-06, showing a fair model fit.

3.  The Influence of Mothers with College Education on Birth Rates The
    coefficient of 0.005423 for mothers with college education is
    notably higher, reflecting a strong positive association with birth
    rates. The standard error is 8.222078e-05, denoting precise
    estimates. The variance of residuals, at 4.577792e-07, suggests a
    good model fit, highlighting the significant impact of higher
    education levels on birth rates.

Conclusion: Overall, these analyses demonstrate varying degrees of
influence that educational attainment and Medicaid coverage have on
birth rates. The positive association with college education is
particularly strong, suggesting that higher education levels may lead to
higher birth rates, perhaps due to increased socio-economic stability.

# **14. Model Fit and** $R^2$ Statistic

The following R2 values represent the proportion of variance in birth
rates explained by each model, highlighting the influence of educational
attainment and Medicaid coverage among mothers:

```{r}
cat("Model 1: Impact of Low Educational Attainment on Birth Rates: ", summary(mlr_nohs)$r.squared, "\n")

cat("Model 2: Influence of Medicaid Coverage on Birth Rates : ", summary(mlr_medi)$r.squared, "\n")

cat("Model 3: Effect of Higher Education on Birth Rates : ", summary(mlr_coll)$r.squared, "\n")

```

Model 1: Impact of Low Educational Attainment on Birth Rates

($R^2$)=0.1822: This indicates that approximately 18.2% of the variance
in birth rates is explained by the proportion of mothers without a high
school diploma. This suggests a moderate impact of low educational
attainment on birth rates.

Model 2: Influence of Medicaid Coverage on Birth Rates

($R^2$)=0.0812: About 8.1% of the variance in birth rates is explained
by Medicaid coverage. This relatively low ($R^2$) suggests a weaker
influence of Medicaid coverage on birth rates compared to educational
factors.

Model 3: Effect of Higher Education on Birth Rates

($R^2$)=0.6402: This model explains 64.0% of the variance in birth
rates, indicating a strong positive relationship between higher
education levels among mothers and birth rates.

The multiple linear regression model, by considering all these factors,
captures more variance than any individual model, providing a
comprehensive understanding of how these socio-economic factors jointly
influence birth rates.

# **15. Discussion**

In this project, I studied how educational attainment and Medicaid
coverage among mothers influence birth rates across different states in
the U.S. The main goal was to understand if these socioeconomic factors
have significant impacts on birth rates and how strong these impacts
might be.

The strongest finding was that mothers with a college education had a
clear positive relationship with birth rates (r=0.8001). This factor
alone explained about 64% of the differences in birth rates. This
suggests that mothers with college education tend to have higher birth
rates, possibly due to better economic stability and resources.

Mothers without a high school diploma also showed a positive but weaker
relationship with birth rates (r=0.4269), explaining about 18.2% of the
variation. Medicaid coverage had the weakest relationship of the three
factors (r=0.2849), explaining only about 8.1%.

When considering all three factors together in a multiple regression
model, I explained around 82.2% of the variation in birth rates, showing
that these factors together significantly influence birth rates.

The study has some limitations. The data used was aggregated at the
state level, which may hide important variations within states, such as
differences between urban and rural areas or between different
socioeconomic groups. Another limitation is the absence of other
potentially influential factors such as income levels, race or
ethnicity, healthcare accessibility, or policy differences.

Future research could use more detailed data, like county or city-level
statistics, and include additional factors. This would allow for a more
complete understanding of what drives birth rates, enabling
better-informed policy decisions and interventions.

Overall, educational attainment, especially college education, has a
significant positive impact on birth rates. Medicaid coverage and lower
educational attainment also influence birth rates but to a lesser
extent.
